---
title: "model_1"
author: "Sam Temlock"
date: "13/11/2020"
output:
  html_document:
    df_print: paged
---

```{r setup, include=FALSE, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(lmtest)
library(sandwich)
library(tidyverse)
library(magrittr)
library(ggplot2)
library(patchwork)
library(stargazer)
library(tinytex)
```

```{r variables, warning=FALSE, results='hide'}
df <- read.csv("covid-19.csv", header = TRUE)
df<-df%>%
  rename(case_rate_100k = 'Case.Rate.per.100000',
         death_rate_100k = 'Death.Rate.per.100000',
         population_density = 'Population.density.per.square.miles',
         white_pct = 'White...of.Cases',
         black_pct = 'Black...of.Cases',
         hispanic_pct = 'Hispanic...of.Cases',
         other_pct = 'Other...of.Cases',
         state_emergency = 'State.of.emergency',
         business_closed = 'Closed.other.non.essential.businesses',
         business_reopen = 'Began.to.reopen.businesses.statewide',
         mask_public='Mandate.face.mask.use.by.all.individuals.in.public.spaces',
         mask_legal='No.legal.enforcement.of.face.mask.mandate') %>%
  select(State, case_rate_100k, death_rate_100k,population_density, white_pct, black_pct, hispanic_pct, 
         other_pct, state_emergency, business_closed, business_reopen, mask_public, mask_legal, 
         Black...of.Total.Population, White...of.Total.Population,
         Percent.living.under.the.federal.poverty.line..2018., Percent.Unemployed..2018., X65.)
# Assign correct data types
cols.num <- c("white_pct","black_pct","hispanic_pct","other_pct")
df[cols.num] <- sapply(df[cols.num],as.numeric)
df$state_emergency=as.Date(df$state_emergency, format = "%m/%d/%Y")
df$business_closed=as.Date(df$business_closed, format = "%m/%d/%Y")
df$business_reopen=as.Date(df$business_reopen, format = "%m/%d/%Y")
head(df)
```

### Model 1

For the first model, the relationship between the COVID case rate per 100,000 and the population density per square mile of states was analyzed. First, the distribution of COVID case rate per 100,000 dependent variable is examined.

```{r}
summary(df$case_rate_100k)
```

```{r}
ggplot(data = df,
  mapping = aes(x= case_rate_100k))+
  geom_histogram(fill = 'skyblue', color = 'grey30', breaks = seq(0,6000,600)) +
  labs(title =  "Histogram of COVID Case Rate per 100,000", x = "Cases per 100,000", y = 'Count')+
  scale_x_continuous(breaks=seq(0, 6000, 600))
```

As can be seen above, the distribution is fairly normal, and given that it has already been standardized as a rate across all states, there is no need to perform any transformations on this variable. Thus, the case rate per 100,000 variable can be leveraged as is as the dependent variable for all three models.

Next, the distribution of population density per square mile variable is examined.

```{r}
histogram_of_pdensity <- df %>%
  ggplot(aes(x = population_density)) + 
  geom_histogram(fill = 'skyblue', color = 'grey30', bins = 20) + 
  labs(
    title = 'Distribution of Population Density per Square Mile', 
    x = 'Population density per square mile', y = 'Count')

histogram_of_pdensity
```

As can be seen from the histogram, although most of the population density is concentrated in the 0 to 1500 range, there are some grouping of outliers that are very far from this concentration. When an analysis is performed, it can be seen that there is only one data sample that is the outlier, which is the District of Columbia (D.C.) with a value of 11,496. This is given due to the fact that D.C. is a district that solely consists of a large city, as mentioned as a possibility in the introduction. Given that this causes the data to be skewed, the logarithm is taken to scale the variable. Alternatively, the data point could have been droppped from the sample, but given the already small sample size, it was determined that a better approach would be to keep it within the sample. Once this transformation was performed, it is shown that there is a relatively normal distribution of population densities (see figure below).

```{r}
# Find the outlier data points
outliers <- subset(df, population_density > 4000)
paste(outliers$State, '=', outliers$population_density)

# Transform the variable by taking the logarithm and assign it to a new variable
df <- df %>% 
  mutate(l_population_density = log(population_density))

# Plot the new distribution in a histogram
histogram_of_pdensity <- df %>%
  ggplot(aes(x = l_population_density)) + 
  geom_histogram(fill = 'skyblue', color = 'grey30', bins = 20) + 
  labs(
    title = 'Figure 2. Distribution of Population Density per Square Mile without D.C.', 
    x = 'Log of population density per square mile', y = 'Count')

histogram_of_pdensity
```

With the appropriate variables transformed, a plot is created to show the relationship between them. 

```{r}
df %>% 
  ggplot(aes(l_population_density, case_rate_100k)) + 
  geom_smooth(se = FALSE) +
  geom_point() +   
  labs(
    title = 'COVID Case Rate due to Population Density', 
    x = 'Log of population density per square mile', 
    y = 'COVID case rate per 100,000'
  )
```

From the above plot, it can be posited that there is no discernible relationship between the variables given the non-linear relationship. In order to test this, the following equation is used to create a regression model to determine the true relationship between the case rate and population density variables.

$$
Case.Rate.Per.100000 =\beta_0 +\beta_1 log(Population.Density.Per.Square.Miles)
$$

```{r}
model1 <- lm(case_rate_100k ~ log(population_density) , data = df)
coeftest(model1, vcov = vcovHC)
```

From the results of the regression, two things are identified. Firstly, the result of the coefficient is nowhere near significant, and therefore we fail to reject the null hypothesis that there is no correlation between case rate and population density, and there is no detectable effect of the independent variable population density. However, the coefficient of population density appears to be negative, which would concur with the assumption that the higher the population density, the lower the COVID case rate in a state.